\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-02 paper.qmd

bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$\ }
  \newcommand\kD{$k\text{-}D$\ }
  \newcommand\dD{$d\text{-}D$\ }
---

@@

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the $k-D$ minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

The array of layouts in @fig-NLDR-variety illustrate what can emerge from the choices of method and parameters, and the random seed that initiates the computation. Key structures interpreted from these views suggest:

- highly **separated clusters** (a, b, e, g, h) with the number ranging from 3-6, 
- **stringy branches** (f), and 
- **barely separated clusters** (c, d) which apparently **contradicts** the other representations.

@@

@tbl-notation summarises the notation used to explain the new methodology. $X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top$ represents the observed data matrix, that are $n$ \pD vectors sampled from some population.

@@

@@

To illustrate the method, we use simulated data. The S-curve is generated as follows XXX Jayani

First, a random variable $a$ is generated from a uniform distribution between $-\frac{\pi}{2}$ and $\frac{\pi}{2}$, representing the parameter $a$ is in $T$. Next, three variables $x_1$, $x_2$, and $x_3$ are computed based on $a$ to construct the S-curve data. $x_1$ is calculated as the sine of $a$ (@eq-eqn1), $x_2$ as twice the value of another uniformly distributed random variable between 0 and 1 (@eq-eqn2), and $x_3$ as the cosine of $a$ minus 1, multiplied by the sign of $a$ (@eq-eqn3). Additionally, four noise dimensions are introduced to the dataset to introduce noise variability: $x_4$, $x_5$, $x_6$, and $x_7$, each generated from uniform distributions within specific intervals (@eq-eqn4, @eq-eqn5, and @eq-eqn6 respectively). Finally, these variables are combined to form the S-curve dataset $T$.

@@

$$
x_7 \sim U(-0.01, 0.01)
$$ {#eq-eqn6}

>>>>

bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$\ }
  \newcommand\kD{$k\text{-}D$\ }
  \newcommand\dD{$d\text{-}D$\ }
---

@@

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the \kD minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

The array of layouts in @fig-NLDR-variety illustrate what can emerge from the choices of method and parameters, and the random seed that initiates the computation. Key structures interpreted from these views suggest: (1) highly **separated clusters** (a, b, e, g, h) with the number ranging from 3-6; (2) **stringy branches** (f), and (3) **barely separated clusters** (c, d) which would **contradict** the other representations.

@@

@tbl-notation summarises the notation used to explain the new methodology. The observed data is denoted as $\mathbfit{x}_{n \times p}$ where $x_{ij}$ would indicate the $i^{th}$ observation on the $j^{th}$ variable sampled from a population $\mathbfit{X}$. To refer to variable $j$, we would use $X_j$.

@@

To illustrate the method, we use $7\text{-}D$ simulated data, which we call the "S-curve". It is constructed by setting $X_1 = \sin(a), X_2 = U(0, 2), X_3 = \text{sign}(a) \times (\cos(a) - 1), \forall a \in [-3\pi/2, 3\pi/2]$. The remaining variables $X_4, X_5, X_6, X_7$ are all uniform error, with small variance. We would consider $T=(X_1, X_2, X_3)$ to be the true model.

@@

<!--First, a random variable $a$ is generated from a uniform distribution between $-\frac{\pi}{2}$ and $\frac{\pi}{2}$, representing the parameter $a$ is in $T$. Next, three variables $x_1$, $x_2$, and $x_3$ are computed based on $a$ to construct the S-curve data. $x_1$ is calculated as the sine of $a$ (@eq-eqn1), $x_2$ as twice the value of another uniformly distributed random variable between 0 and 1 (@eq-eqn2), and $x_3$ as the cosine of $a$ minus 1, multiplied by the sign of $a$ (@eq-eqn3). Additionally, four noise dimensions are introduced to the dataset to introduce noise variability: $x_4$, $x_5$, $x_6$, and $x_7$, each generated from uniform distributions within specific intervals (@eq-eqn4, @eq-eqn5, and @eq-eqn6 respectively). Finally, these variables are combined to form the S-curve dataset $T$.

@@

$$
x_7 \sim U(-0.01, 0.01)
$$ {#eq-eqn6}
-->

### 2024-04-02 paper.tex

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$\ }
\newcommand\kD{$k\text{-}D$\ }
\newcommand\dD{$d\text{-}D$\ }
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

Many new methods for NLDR have emerged in recent years, all designed to
better capture specific structures potentially existing in \pD. Here we
focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP
and PaCMAP. tNSE and UMAP can be considered to produce the \(k-D\)
minimizing the divergence between two distributions, where the
distributions are modeling the inter-point distances. PHATE, TriMAP and
PaCMAP are examples of diffusion processes \citep{coifman2005} spreading
to capture geometric shapes, that include both global and local
structure.

The array of layouts in Figure~\ref{fig-NLDR-variety} illustrate what
can emerge from the choices of method and parameters, and the random
seed that initiates the computation. Key structures interpreted from
these views suggest:

\begin{itemize}
\tightlist
\item
  highly \textbf{separated clusters} (a, b, e, g, h) with the number
  ranging from 3-6,
\item
  \textbf{stringy branches} (f), and
\item
  \textbf{barely separated clusters} (c, d) which apparently
  \textbf{contradicts} the other representations.
\end{itemize}

@@

Table~\ref{tbl-notation} summarises the notation used to explain the new
methodology.
\(X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top\)
represents the observed data matrix, that are \(n\) \pD vectors sampled
from some population.

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$X$ & $p$-dimensional data\\
$Y$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d-dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$y_{max}$ & maximum value of the scaled second embedding component\\
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
\addlinespace
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
\addlinespace
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C^{(j)}$ & $j$-dimensional bin centers\\
\bottomrule
\end{tabular}
\endgroup{}

@@

@@

To illustrate the method, we use simulated data. The S-curve is
generated as follows XXX Jayani

First, a random variable \(a\) is generated from a uniform distribution
between \(-\frac{\pi}{2}\) and \(\frac{\pi}{2}\), representing the
parameter \(a\) is in \(T\). Next, three variables \(x_1\), \(x_2\), and
\(x_3\) are computed based on \(a\) to construct the S-curve data.
\(x_1\) is calculated as the sine of \(a\) (Equation~\ref{eq-eqn1}),
\(x_2\) as twice the value of another uniformly distributed random
variable between 0 and 1 (Equation~\ref{eq-eqn2}), and \(x_3\) as the
cosine of \(a\) minus 1, multiplied by the sign of \(a\)
(Equation~\ref{eq-eqn3}). Additionally, four noise dimensions are
introduced to the dataset to introduce noise variability: \(x_4\),
\(x_5\), \(x_6\), and \(x_7\), each generated from uniform distributions
within specific intervals (Equation~\ref{eq-eqn4},
Equation~\ref{eq-eqn5}, and Equation~\ref{eq-eqn6} respectively).
Finally, these variables are combined to form the S-curve dataset \(T\).

\begin{equation}\phantomsection\label{eq-eqn1}{
x_1 = \sin(3 \pi a)
}\end{equation}

\begin{equation}\phantomsection\label{eq-eqn2}{
x_2 = 2.0 \times U(0, 1)
}\end{equation}

\begin{equation}\phantomsection\label{eq-eqn3}{
x_3 = \text{sign}(a) \times (\cos(3 \pi a) - 1)
}\end{equation}

\begin{equation}\phantomsection\label{eq-eqn4}{
x_4, x_5 \sim U(-0.02, 0.02)
}\end{equation}

\begin{equation}\phantomsection\label{eq-eqn5}{
x_6 \sim U(-0.1, 0.1)
}\end{equation}

\begin{equation}\phantomsection\label{eq-eqn6}{
x_7 \sim U(-0.01, 0.01)
}\end{equation}

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$\ }
\newcommand\kD{$k\text{-}D$\ }
\newcommand\dD{$d\text{-}D$\ }
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

Many new methods for NLDR have emerged in recent years, all designed to
better capture specific structures potentially existing in \pD. Here we
focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP
and PaCMAP. tNSE and UMAP can be considered to produce the
\kD minimizing the divergence between two distributions, where the
distributions are modeling the inter-point distances. PHATE, TriMAP and
PaCMAP are examples of diffusion processes \citep{coifman2005} spreading
to capture geometric shapes, that include both global and local
structure.

The array of layouts in Figure~\ref{fig-NLDR-variety} illustrate what
can emerge from the choices of method and parameters, and the random
seed that initiates the computation. Key structures interpreted from
these views suggest: (1) highly \textbf{separated clusters} (a, b, e, g,
h) with the number ranging from 3-6; (2) \textbf{stringy branches} (f),
and (3) \textbf{barely separated clusters} (c, d) which would
\textbf{contradict} the other representations.

@@

Table~\ref{tbl-notation} summarises the notation used to explain the new
methodology. The observed data is denoted as
\(\mathbfit{x}_{n \times p}\) where \(x_{ij}\) would indicate the
\(i^{th}\) observation on the \(j^{th}\) variable sampled from a
population \(\mathbfit{X}\). To refer to variable \(j\), we would use
\(X_j\).

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$\mathbfit{X}, \mathbfit{x}$ & $p$-dimensional data (population, sample)\\
$Y$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d\text{-}dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$y_{max}$ & maximum value of the scaled second embedding component\\
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
\addlinespace
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
\addlinespace
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C^{(j)}$ & $j$-dimensional bin centers\\
\bottomrule
\end{tabular}
\endgroup{}

@@

To illustrate the method, we use \(7\text{-}D\) simulated data, which we
call the ``S-curve''. It is constructed by setting
\(X_1 = \sin(a), X_2 = U(0, 2), X_3 = \text{sign}(a) \times (\cos(a) - 1), \forall a \in [-3\pi/2, 3\pi/2]\).
The remaining variables \(X_4, X_5, X_6, X_7\) are all uniform error,
with small variance. We would consider \(T=(X_1, X_2, X_3)\) to be the
true model.

@@
